\section{Scope}

\npf This document focuses only on the UPC specifications
    that extend the C Standard to an explicit parallel C based on the
    partitioned global address space model. All C specifications as per
    ISO/IEC 9899 [ISO/IEC00] are considered a part of these UPC
    specifications, and therefore will not be addressed in this
    document.
\index{ISO C}
\index{UPC}
\index{global address space}

\np Small parts of the C Standard [ISO/IEC00] may be repeated for
    self-containment and clarity of a subsequent UPC extension
    definition.

\section{Normative references}

\xchangenote[id=DB]{100}{ADDED [UPC-LIB-REQ] and [UPC-LIB-OPT]}

\npf The following documents and their identified normative
     references constitute provisions of these UPC specifications.
     For dated references, only the edition cited applies. For undated references,
     the latest edition of the referenced document applies.

\index{ISO C}
\np ISO/IEC 9899: 1999(E), Programming languages - C
    [ISO/IEC00]

\index{UPC Required Library}
\np UPC Required Library Specifications
    [UPC-LIB-REQ]

\index{UPC Optional Library}
\np UPC Optional Library Specifications
    [UPC-LIB-OPT]

\np The relationship between the section numbering used in
    the C Standard [ISO/IEC00] and that used in this document is given
    in Appendix C and noted at the beginning of each corresponding section.

\np Implementations shall document the exact revisions of 
    [UPC-LIB-REQ] and [UPC-LIB-OPT] to which they conform.

