\subsection{Castability Functions \texttt{<}upc\_castable.h\texttt{>}}
\label{upc-castable}
\index{privatizability}
\index{castability}
\index{upc\_castable.h}
\index{\_\_UPC\_CASTABLE\_\_}

\npf A UPC implementation may map some or all of the shared address space
of another thread into the local address space of the current thread.  The
functions described in this section allow the programmer to determine
whether this is the case, and to make use of this information by
providing the ability to obtain valid local addresses for shared cells
with affinity to other threads.  This capability, sometimes called
"privatizability", is referred to as "castability" in this section.

\np Implementations that support this interface shall predefine the
feature macro {\_\_UPC\_CASTABLE\_\_} to the value 1.

\subsubsection{Standard headers}

\npf The standard header is

{\tt <upc\_castable.h>}

\np Unless otherwise noted, all of the functions, types and macros
specified in Section~\ref{upc-castable}
are declared by the header {\tt <upc\_castable.h>}.

\subsubsection{Castability Functions}

\paragraph{The {\tt upc\_cast} function}
\index{upc\_cast}
\index{cast}

{\bf Synopsis}

\npf\vspace{-2.5em}
\begin{verbatim}
    #include <upc_castable.h>
    void *upc_cast(const shared void *ptr);
\end{verbatim}

{\bf Description}

\np The {\tt upc\_cast} function converts the specified
pointer-to-shared to a valid pointer-to-local.  If such a conversion
is not possible, a null pointer is returned.

\np The pointer {\tt ptr} points into one or more shared objects.
Consider the portions of all of these shared objects with affinity to
{\tt upc\_threadof(ptr)}.  Choose the shared object
containing {\tt ptr} where the portion with this affinity is largest.
The conversion performed by the
{\tt upc\_cast} function will be considered possible only if this entire
portion may be read and written by the current thread based on the returned
pointer-to-local value.

\index{castable}
\np If the conversion is possible, the pointer-to-shared value is referred to as
{\em castable}.

\np If {\tt upc\_threadof(ptr)} is equal to
{\tt MYTHREAD}, {\tt upc\_cast(ptr)} is equivalent to {\tt (void *)ptr}.

\np If the {\tt ptr} pointer is null, {\tt upc\_cast} returns a null pointer.

\np The pointer returned by {\tt upc\_cast} is valid only in the calling
thread.  It cannot be assumed that the return value may be passed to
a different thread and used by that thread.  It also cannot be assumed that two
threads calling {\tt upc\_cast} with the same argument will get the same
return value.

\np The pointer returned by {\tt upc\_cast} remains valid for the lifetime
of the referenced shared object.  In particular, if the referenced shared
object was dynamically allocated, the pointer is no longer valid after the 
associated shared memory has been freed.

\np If a call to {\tt upc\_cast} succeeds, subsequent calls by the same thread
with the same pointer, or with a pointer into the same object and with the same
affinity, are also guaranteed to succeed for the lifetime of the object.

\paragraph{The {\tt upc\_thread\_info} function}
\index{upc\_thread\_info}

{\bf Synopsis}

\npf\vspace{-2.5em}
\begin{verbatim}
    #include <upc_castable.h>
    upc_thread_info_t upc_thread_info(size_t threadId);
\end{verbatim}

{\bf Description}

\index{upc\_thread\_info\_t}
\np The {\tt upc\_thread\_info} function returns information about
potential uses of the {\tt upc\_cast} function in the calling
thread in reference to objects with
affinity to thread {\tt threadId}.  The information is returned in a
{\tt upc\_thread\_info\_t} structure, with the following fields:

\begin{description}
\item{\tt int guaranteedCastable} \hfill \\
Indicates which memory regions are guaranteed to be castable.
\item{\tt int probablyCastable} \hfill \\
Indicates which memory regions are likely (but not guaranteed)
to be castable.
\end{description}

\np An implementation may provide additional fields in this structure,
allowing {\tt upc\_thread\_info} to return other information about
thread {\tt threadId} with respect to the calling thread.

\np The {\tt guaranteedCastable} and {\tt probablyCastable} fields
contain coded integer values indicating memory regions.  If the flag
for a particular region is set in the {\tt guaranteedCastable} field,
it indicates that
any pointer into that region with affinity to {\tt threadId} is
castable.  If the flag is
set for a particular region is set in the {\tt probablyCastable} field,
it indicates that it is likely, but not guaranteed, that a pointer into
that region with affinity to {\tt threadId} is castable.

\np Flag values are as specified below.  Each flag has a
unique bitwise representation that can be unambiguously tested using the
bitwise AND operator (\&).

\begin{description}
\index{UPC\_CASTABLE\_ALL\_ALLOC}
\item[{\tt UPC\_CASTABLE\_ALL\_ALLOC}] \hfill \\
Refers to memory allocated via {\tt upc\_all\_alloc}.
\index{UPC\_CASTABLE\_GLOBAL\_ALLOC}
\item[{\tt UPC\_CASTABLE\_GLOBAL\_ALLOC}] \hfill \\
Refers to memory allocated via {\tt upc\_global\_alloc}.
\index{UPC\_CASTABLE\_ALLOC}
\item[{\tt UPC\_CASTABLE\_ALLOC}] \hfill \\
Refers to memory allocated via {\tt upc\_alloc}.
\index{UPC\_CASTABLE\_STATIC}
\item[{\tt UPC\_CASTABLE\_STATIC}] \hfill \\
Refers to shared variables defined via static and file scope declarations.
\end{description}

\np Implementations may define additional memory region flags.

\np The macro {\tt UPC\_CASTABLE\_ALL} shall
be defined to be all the region-specific values (including any
implementation-specific values) combined via bitwise
OR ({\textbar}) operations.  It is defined for convenient testing of
whether all shared memory regions are covered in the returned flag.

\np If no memory regions are indicated by the returned flag, the flag
value shall be zero.

